---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Yiting Zhong"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **CA Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)
 
# Set your Census API key
census_api_key("577f52cafa8132078e715d2f9f7321573a862bd1")

# Choose your state for analysis - assign it to a variable called my_state
my_state="CA"
```

**State Selection:** I have chosen **CA** for this analysis because: it has diverse culture.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
county_data <- get_acs(
  geography = "county",
  variables = c(
    med_income = "B19013_001",
    total_pop = "B01003_001"
  ),
  state = my_state,
  year = 2022,
  survey = "acs5",
  output = "wide"
)

# Clean the county names to remove state name and "County" 

# Hint: use mutate() with str_remove()
county_data_clean <- county_data %>%
  mutate(NAME = str_remove(NAME, "County, California"))

# Display the first few rows
head(county_data_clean)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
income_analysis <- county_data_clean %>%
  mutate(
    moe_pct = (med_incomeM / med_incomeE) * 100,
    reliability = case_when(
      moe_pct < 5 ~ "High Confidence",
      moe_pct <= 10 ~ "Moderate Confidence",
      TRUE ~ "Low Confidence"
    ),
    unreliable_flag = if_else(moe_pct > 10, "Yes", "No")
  )

# Create a summary showing count of counties in each reliability category
reliability_summary <- income_analysis %>%
  count(reliability) %>%
  mutate(
    percentage = (n / sum(n)) * 100
  )

kable(reliability_summary, digits = 1, caption = "Summary of Data Reliability by County")

# Hint: use count() and mutate() to add percentages
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
high_uncertainty_top5 <- income_analysis %>%
  arrange(desc(moe_pct))%>%
  slice(1:5)%>%
  select(
    NAME,
    med_incomeE,
    med_incomeM,
    moe_pct,
    reliability
  )

# Format as table with kable() - include appropriate column names and caption
kable(
high_uncertainty_top5, 
      col.names = c("County Name", "Median Income (Estimate)", "Margin of Error", "MOE %", "Reliability"),
digits = 2,
caption = "Top 5 counties with the highest Uncertainty"
)
```

**Data Quality Commentary:**

[Write 2-3 sentences explaining what these results mean for algorithmic decision-making. Consider: Which counties might be poorly served by algorithms that rely on this income data? What factors might contribute to higher uncertainty?]

"The high MOE percentages in rural counties like Mono and Alpine (both over 17%) indicate that small sample sizes in sparsely populated areas lead to significant data instability. If algorithmic systems treat these "Low Confidence" estimates as absolute facts, they risk misallocating social services and reinforcing geographic inequities against rural communities. Consequently, these communities may be poorly served by automated decisions that fail to account for the inherent statistical uncertainty in their income data."

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_counties<- income_analysis%>%
  group_by(reliability) %>%
  slice(1) %>%
  ungroup() %>%
  select(NAME, GEOID, med_incomeE, moe_pct, reliability)

# Display the selected counties with their key characteristics
kable(
selected_counties, 
      col.names = c("County Name", "GEOID", "Median Income (Estimate)", "MOE %", "Reliability"),
digits = 3,
caption = "Representative Counties Selected for Each Reliability Level"
)
# Show: county name, median income, MOE percentage, reliability category
```

**Comment on the output:** [I selected "Alameda", ""Amador", and "Alpine" to represent High, Moderate, and Low Confidence levels respectively. ]

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
race_var <- c(
  total_pop= "B03002_001",
  white_alone="B03002_003",
  black="B03002_004",
  hispanic_latino="B03002_012"
)
# Use get_acs() to retrieve tract-level data
target_counties <- str_sub(selected_counties$GEOID, 3, 5)

tract_data<-get_acs(
  geography = "tract",
  variables = race_var,
  state = my_state,
  county = target_counties,
  year = 2022,
  survey = "acs5",
  output = "wide"
  
)
# Hint: You may need to specify county codes in the county parameter

# Calculate percentage of each group using mutate()
tract_demographics <- tract_data %>%
  mutate(
    pct_white= (white_aloneE/total_popE)*100,
    pct_black=(blackE/total_popE)*100,
    pct_h_l=(hispanic_latinoE/total_popE)*100,
    county_name = str_extract(NAME, "(?<=; ).*(?= County;)"),
    tract_name = str_extract(NAME, "^[^,]*"
  ))
# Create percentages for white, Black, and Hispanic populations

# Add readable tract and county name columns using str_extract() or similar
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
highest_h_l <- tract_demographics %>%
  arrange(desc(pct_h_l)) %>%
  slice(1) %>%
  select(NAME,pct_h_l, total_popE )
# Calculate average demographics by county using group_by() and summarize()
average_demographics_couty <- tract_demographics %>%
  group_by(county_name) %>%
  summarise(
    tract_count= n(),
    avg_white=mean(pct_white, na.rm = TRUE),
    avg_black=mean(pct_black, na.rm = TRUE),
    avg_h_l=mean( pct_h_l, na.rm = TRUE),
  )
# Show: number of tracts, average percentage for each racial/ethnic group

# Create a nicely formatted table of your results using kable()
kable(
average_demographics_couty, 
      col.names = c("County", "Total Tracts", "Avg White %", "Avg Black %", "Avg Hispanic %"),
digits = 3,
caption = "Demographic Analysis of Selected County"
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
moe_analysis <- tract_data %>%
  mutate(
  moe_white = ((white_aloneM/white_aloneE)*100),
  moe_black = ((blackM/blackE)*100),
  moe_h_l = ((hispanic_latinoM/hispanic_latinoE)*100),
  moe_total_pop=((total_popM/total_popE)*100),
  high_moe_flag = if_else(
  (is.infinite(moe_white)& moe_white >15)|
  (is.infinite(moe_black)& moe_black >15)|
  (is.infinite(moe_h_l)& moe_h_l >15),
  "High Uncertainty","Reliable"
))

# Create summary statistics showing how many tracts have data quality issues
moe_sum<- moe_analysis %>%
  count(high_moe_flag) %>%
  mutate(pct =(n/sum(n))*100)

kable(moe_sum, caption = "Summary of Tract-Level Data Quality Issues")
  
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison

patten_analysis <- moe_analysis %>%
  mutate(
    pct_white=((white_aloneE/total_popE)*100),
    pct_black=((blackE/total_popE)*100),
    pct_h_l=((hispanic_latinoE/total_popE)*100)
  )%>%
  group_by(high_moe_flag)%>%
  summarise(
    avg_white=mean(pct_white, na.rm = TRUE),
    avg_black=mean(pct_black, na.rm = TRUE),
    avg_h_l=mean(pct_h_l, na.rm = TRUE),
    avg_total=mean(total_popE, na.rm = TRUE),
    tract_count=n()
  )
# Create a professional table showing the patterns
kable(patten_analysis,
      col.names = c("Data Quality Group", "Avg white %", "Avg Black %", "Avg Hispanic %","Avg Total Pop", "Tract Counts"),
digits = 2,
caption = "Summary of Pattern Analysis")
```

**Pattern Analysis:** [Describe any patterns you observe. Do certain types of communities have less reliable data? What might explain this?]
Tracts with sparse populations exhibit high uncertainty due to limited sample sizes, resulting in imprecise data and a significant margin of error. In analyses, data from white populations in high-uncertainty tracts appear unreliable not because of racial issues, but because residents in sparsely populated rural areas tend to be more ethnically homogeneous than those in diverse urban centers. Consequently, higher average proportions of white residents are observed within the “high uncertainty” group.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

[Your integrated 4-paragraph summary here]
1. Across all four levels of analysis, a clear systematic pattern emerges: data reliability is inversely correlated with population density. At the county level, urban centers like Alameda exhibit high confidence with a margin of error (MOE) as low as 1.005%, whereas sparsely populated rural counties like Alpine suffer from low confidence with MOEs exceeding 17.2%. This trend persists at the tract level, where "High Uncertainty" areas have a significantly smaller average population (2,729) compared to "Reliable" tracts (4,435).

2. The communities facing the greatest risk of algorithmic bias are small, rural, and geographically isolated populations. Based on the pattern analysis, these "High Uncertainty" tracts often have a higher average percentage of White residents (48.83%) and significantly lower percentages of Black residents (0.83%) compared to urban centers. While urban minority communities benefit from the robust data of high-density areas, rural residents—regardless of race—are at risk of being "miscalculated" by algorithms that do not account for the high statistical noise inherent in small-sample data.

3. The primary driver of this bias risk is the inherent limitation of sampling in low-density areas. In a county like Alpine, which has a total population of only 1,515, the American Community Survey (ACS) sample size is so small that a few outlier responses can drastically skew the median income estimate. Algorithms treat these estimates as absolute facts, but in reality, they are "statistical flukes" caused by the geographic reality of rural life, creating a structural bias where data-rich urban areas receive more predictable and "fair" algorithmic treatment than data-sparse rural ones.

4. To mitigate these risks, the Department of Human Services should implement a "Reliability-Aware" Decision Framework. First, any community flagged with an MOE > 10% should be diverted from fully automated processing to a manual administrative review. Second, the Department should supplement Census data with local administrative records (such as state tax or utility data) for rural tracts to triangulate income estimates. 


## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

decision_framework <- income_analysis %>%
  select( NAME, med_incomeE, moe_pct,total_popE,reliability,) %>%
  mutate(
    algo_recommendation = case_when(
      reliability == "High Confidence"~"Safe for algorithmic decisions",
      reliability == "Moderate Confidence"~"Use with caution - monitor outcomes",
      reliability == "Low Confidence"~"Requires manual review or additional data"
  ))
# Format as a professional table with kable()
kable(decision_framework,
      col.names = c("county name", "median income", "MOE percentage", "Total population","Reliability","Policy Recommendation"),
digits = 2,
caption = "Decision Framework for Algorithm Implementation")
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** [List counties with high confidence data and explain why they're appropriate]
Alameda 
Butte 
Calaveras 
Contra Costa 
El Dorado 
Fresno 
Humboldt 
Imperial 
Kern 
Kings 
Lake 
Los Angeles 
Madera 
Marin 
Mendocino 
Merced 
Monterey 
Napa 
Nevada 
Orange 
Placer 
Riverside 
Sacramento 
San Bernardino 
San Diego 
San Francisco 
San Joaquin 
San Luis Obispo 
San Mateo 
Santa Barbara 
Santa Clara 
Santa Cruz 
Shasta 
Siskiyou 
Solano 
Sonoma 
Stanislaus 
Sutter 
Tulare 
Ventura 
Yolo 
Yuba 
These counties have large populations and exceptionally large sample sizes, resulting in an extremely low margin of error (MOE) of approximately 1%. Their income estimates are highly stable, and the risk of algorithmic bias in these areas is minimal, making them safe for automated funding allocation.

2. **Counties requiring additional oversight:** [List counties with moderate confidence data and describe what kind of monitoring would be needed]
Amador 
Colusa 
Del Norte 
Glenn 
Inyo 
Lassen 
Mariposa 
Modoc 
San Benito 
Tehama 
Tuolumne 
Although the data is generally accurate, it exhibits a certain degree of volatility. It is recommended that while implementing the algorithm, an output monitoring mechanism be established to periodically conduct random sampling and inspection of anomalous allocation cases, ensuring the algorithm does not develop bias due to data fluctuations.

3. **Counties needing alternative approaches:** [List counties with low confidence data and suggest specific alternatives - manual review, additional surveys, etc.]
Alpine 
Mono 
Plumas 
Sierra 
Trinity 
These counties have sparse populations, resulting in extremely unstable data with errors exceeding 17%. Direct application of algorithms would lead to severe geographic inequities. A “manual verification” mechanism could be implemented for these counties, or supplementary investigations could be conducted using state government administrative records to compensate for the shortcomings of census sample data.

## Questions for Further Investigation

[List 2-3 questions that your analysis raised that you'd like to explore further in future assignments. Consider questions about spatial patterns, time trends, or other demographic factors.]
1.Has the reliability of data for these remote counties improved or deteriorated over the past decade as the Census Bureau updated its sampling methods?
2.Could incorporating “income” alongside other metrics help reduce bias in the final results?
# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on February 9, 2026.

**Reproducibility:** 
- All analysis conducted in R version 4.3.0
- Census API key required for replication
- Complete code and documentation available at: [your portfolio URL]

**Methodology Notes:**
County and tract names were cleaned using regular expressions to remove repetitive legal suffixes (e.g., "; California"), ensuring the final visualizations and tables were policy-ready.

**Limitations:**
The focus was restricted to California (CA); findings regarding rural data uncertainty may vary in states with different administrative structures or population distributions.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`