---
title: "Bikeshare_SpaceTime_Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Yiting Zhong"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: **rebalancing bikes to meet anticipated demand**.

Imagine you're an Indego operations manager at 6:00 AM on a Monday morning. You have: - 200 stations across Philadelphia - Limited trucks and staff for moving bikes - 2-3 hours before morning rush hour demand peaks - **The question:** Which stations will run out of bikes by 8:30 AM?

This lab will teach you to build predictive models that forecast bike share demand across **space** (different stations) and **time** (different hours) to help solve this operational problem.

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)

library(knitr)
library(patchwork)
library(kableExtra)
library(here)

options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}

census_api_key(Sys.getenv("CENSUS_API_KEY"))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

# IMPORT AND PREP DATA

## Load Indego Trip Data (Q2 2025)

**Note for HW5:** You'll download a different quarter from: https://www.rideindego.com/about/data/

```{r load_indego}
# Read Q1 2025 data
indego <- read_csv(here("C:/Users/yzhon/OneDrive - PennO365/Penn-Spring 2026/CPLN 592/cpln-5920-sp26-student-portfolio-template/Labs/Lab_5/indego-trips-2025-q2.csv"))

# Quick look at the data
glimpse(indego)
```

**Pause here** Look at the data. What do you suspect the coordinate system is? What do you notice about the date and time structure?

> The coordinate fields suggest that the data uses geographic coordinates in latitude and longitude, likely in decimal degrees (such as WGS84). The time variables are stored as full timestamps with both date and time, rather than date alone. Because each trip has a start and end timestamp, the data will likely need to be aggregated to hourly intervals for station-hour analysis.

## Examine the Data Structure

```{r explore_data}
# How many trips?
cat("Total trips in Q2 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure. Read up on the lubridate library for more information on the date/time data extraction.

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

**Question** Explain the interval60 feature to someone else at your table.

> `interval60` is an hourly time bin created from the trip start time. It rounds each trip down to the start of the hour, so trips that begin within the same hour are grouped together. This is useful because the analysis predicts bike demand at the station-hour level rather than for individual trips.

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2025",
    subtitle = "Spring to early summer demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

**Question:** What patterns do you see? How does ridership change over time?

The plot suggests that Indego ridership rises overall during Q2 2025. The red smoothed line increases from April through June, indicating stronger demand as the quarter progresses. In addition, the blue daily series shows regular up-and-down variation, which suggests a repeating weekly cycle in bike use. There are also several unusually low-ridership days, which may reflect weekends, weather conditions, or other one-off disruptions.

```         
```

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

**Question:** When are the peak hours? How do weekends differ from weekdays?

The busiest hours on weekdays are around 8 AM and especially 4–6 PM, with the strongest peak at 5 PM. In contrast, weekends do not show a sharp morning peak. Instead, weekend ridership builds gradually and remains high from around noon through the late afternoon. This suggests a commuter pattern on weekdays and a more leisure-oriented pattern on weekends.

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

We'll get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_gradient(
    low = "lightyellow",
    high = "darkgreen",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

**Question** These data are in WGS84. If you wanted to calculate a spatial distance-based variable, like a buffer, should you use the dataset as it?

No. WGS84 stores coordinates in latitude and longitude (degrees), so it is not ideal for calculating distance-based measures such as buffers. For spatial distance calculations, the data should first be projected into a coordinate system with linear units such as meters or feet.

## Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_gradient(
    low = "lightyellow",
    high = "darkgreen",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

## Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations..

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

## Get Weather Data

Weather significantly affects bike share demand! Let's get hourly weather for Philadelphia.

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
library(riem)
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-04-01",
  date_end = "2025-06-30"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

```{r visualize-weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2025",
    subtitle = "Spring to early summer transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

**Note:** We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

Temporal lags: good Spatial lags: bad!

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# TEMPORAL TRAIN/TEST SPLIT

**CRITICAL:** We must train on PAST data and test on FUTURE data!

## Why Temporal Validation Matters

In real operations, at 6:00 AM on March 15, we need to predict demand for March 15-31. We have data from Jan 1 - March 14, but NOT from March 15-31 (it hasn't happened yet!).

**Wrong approach:** Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

**Correct approach:** Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r temporal-split}
early_stations <- study_panel_complete %>%
  filter(date < as.Date("2025-06-01")) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(date >= as.Date("2025-06-01")) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

common_stations <- intersect(early_stations, late_stations)

study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

train <- study_panel_complete %>%
  filter(date < as.Date("2025-06-01"))

test <- study_panel_complete %>%
  filter(date >= as.Date("2025-06-01"))

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

#BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

-   Tuesday and Thursday have positive coefficients (**+0.031** and **+0.062**), while Wednesday is almost the same as Monday (**+0.002**)
-   Friday is slightly below Monday (**-0.016**)
-   This suggests weekday ridership is fairly stable, but Thursday appears strongest and Friday is slightly weaker

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (**-0.036** for Saturday and **-0.087** for Sunday)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Overnight hours are the lowest, with hours 1–4 all below the midnight baseline. Ridership rises quickly after 6 AM, reaches a strong morning peak at 8 AM (+0.847), and stays elevated through midday. The strongest peak occurs at 5 PM (+1.188), showing an even stronger evening commute than morning commute.

## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

**Question:** Did adding lags improve R²? Why or why not?

Yes, adding lag variables improved the model substantially. The R² increased from **0.112** in Model 1 to **0.309** in Model 2, which means the model explains much more variation in station-hour trips than before. This improvement makes sense because bike-share demand is strongly related to recent past demand. If a station was busy one hour ago, three hours ago, or at the same hour the previous day, it is more likely to be busy again now. These lag variables capture temporal persistence and repeated travel patterns, especially commuting behavior.

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

**What do station fixed effects capture?**

Station fixed effects capture persistent baseline differences across stations. In other words, some stations are simply busier than others because of their location and surrounding context, such as land use, transit connections, and neighborhood activity.

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question:** Which features gave us the biggest improvement?

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

**Question:** Where is the model performing well? Where is it struggling?

-   The model performs **best at low to moderate trip counts**, where the points are more concentrated and the fitted green line follows the overall trend reasonably well

<!-- -->

-   It seems to do **relatively better overnight and during mid-day**, when demand is lower and less variable

<!-- -->

-   The model struggles more during **PM Rush**, where there are many high observed trip counts but the predicted values stay too low

<!-- -->

-   This means the model tends to **underpredict high-demand periods**, especially during rush hours

<!-- -->

-   In general, the farther the observed values move away from zero, the more the predictions flatten out below the red perfect-prediction line

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors}
# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_gradientn(
    colours = c("lightyellow", "orange", "red", "darkred"),
    name = "MAE (trips)",
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_gradientn(
    colours = c("lightblue", "deepskyblue3", "blue", "darkblue"),
    name = "Avg Demand (trips/hour)",
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

**Question:** Do you see spatial clustering of errors? What neighborhoods have high errors?

Yes, the errors appear to be spatially clustered rather than randomly distributed. The highest prediction errors are concentrated in Center City and nearby parts of West Philadelphia. These same areas also show relatively high average demand, suggesting that the model has the most difficulty in the busiest parts of the network. Outer stations appear to have both lower demand and lower error.

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

**Critical Question:** Are prediction errors systematically higher in certain demographic groups? What are the equity implications?

The scatterplots suggest that prediction errors may be systematically related to neighborhood characteristics. MAE appears to increase with median income and with the percentage of White residents, while it decreases as transit usage increases. This means the model may be less accurate in some demographic contexts than others, rather than making equally sized errors everywhere. From an equity perspective, this matters because uneven model performance could lead to better service planning in some neighborhoods and worse planning in others.

# 

## **Compare results** to Q1 2025:

1.  Compared with the original Q1 template, Q2 appears to have higher overall ridership, which is consistent with warmer spring and early summer conditions. The same broad commute structure remains visible, with strong morning and evening peaks, but Q2 demand also appears more elevated across midday and afternoon periods. These seasonal differences may affect both model fit and prediction error, since warmer-weather travel can be more frequent and potentially more variable than winter ridership.

## Part 3: Feature Engineering & model improvement

-   Does this improve model fit?

```{r}
# 1. Add new features to the full panel
study_panel_fe <- study_panel_complete %>%
  arrange(start_station, interval60) %>%
  group_by(start_station) %>%
  mutate(
    # Nice biking weather indicator
    nice_weather = ifelse(Temperature >= 60 & Temperature <= 75 & Precipitation == 0, 1, 0),
    
    # Weekend + nice weather interaction
    weekend_nice_weather = ifelse(weekend == 1 & nice_weather == 1, 1, 0),
    
    # Same station, same approximate hour one week earlier
    lag1week = lag(Trip_Count, 168)
  ) %>%
  ungroup()

# 2. Re-create train/test split for Q2
train_fe <- study_panel_fe %>%
  filter(date < as.Date("2025-06-01"))

test_fe <- study_panel_fe %>%
  filter(date >= as.Date("2025-06-01"))

# 3. Re-create day-of-week factor
train_fe <- train_fe %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

test_fe <- test_fe %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# 4. Keep only rows with complete lag information
train_fe <- train_fe %>%
  drop_na(lag1Hour, lag3Hours, lag1day, lag1week)

test_fe <- test_fe %>%
  drop_na(lag1Hour, lag3Hours, lag1day, lag1week)

# 5. Fit improved linear model
model_improved <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + lag1week +
    nice_weather + weekend_nice_weather +
    as.factor(start_station),
  data = train_fe
)


# 6. Predict on test set
test_fe <- test_fe %>%
  mutate(
    pred_improved = predict(model_improved, newdata = test_fe)
  )

# 7. Calculate MAE for improved model
mae_improved <- mean(abs(test_fe$Trip_Count - test_fe$pred_improved), na.rm = TRUE)

cat("Improved Model MAE:", round(mae_improved, 4), "\n")

# 8. Fit Poisson model for count data
poisson_model <- glm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + lag1week +
    nice_weather + weekend_nice_weather,
  family = poisson(link = "log"),
  data = train_fe
)


# 9. Predict from Poisson model
test_fe <- test_fe %>%
  mutate(
    pred_poisson = predict(poisson_model, newdata = test_fe, type = "response")
  )

# 10. Calculate MAE for Poisson model
mae_poisson <- mean(abs(test_fe$Trip_Count - test_fe$pred_poisson), na.rm = TRUE)

cat("Poisson Model MAE:", round(mae_poisson, 4), "\n")

best_old_mae <- 0.83

comparison_table <- tibble(
  Model = c("Best Previous Model", "Improved Linear Model", "Poisson Model"),
  MAE = c(best_old_mae, mae_improved, mae_poisson)
)

print(comparison_table)

cat("\nRows in train_fe:", nrow(train_fe), "\n")
cat("Rows in test_fe:", nrow(test_fe), "\n")
cat("Train date range:", min(train_fe$date), "to", max(train_fe$date), "\n")
cat("Test date range:", min(test_fe$date), "to", max(test_fe$date), "\n")
```

I added three new features based on the earlier error analysis:

**`nice_weather`** : an indicator for favorable biking weather, defined as temperatures between 60°F and 75°F with no precipitation

**`weekend_nice_weathe`**r: an interaction term to capture the idea that pleasant weather may increase leisure riding on weekends

**`lag1week`**: the trip count from the same station one week earlier, which helps capture repeated weekly demand patterns

These features were motivated by the patterns found in the earlier analysis. The temporal plots showed strong repeated demand cycles, which suggested that past demand from the same time last week could be useful. The Q2 context also made weather especially important, since spring and early summer biking is likely to be sensitive to pleasant conditions. In addition, weekend demand may be shaped by recreational travel, so a weekend-by-weather interaction was a reasonable extension.

The improved linear model reduced MAE from 0.830 in the previous best model to 0.675, which is a substantial improvement. This suggests that the new features captured important patterns that were missing from the earlier models.

I also estimated a Poisson model because trip counts are count data. The Poisson model achieved an MAE of 0.723, which was better than the previous best model but still worse than the improved linear model. This suggests that although the Poisson framework is conceptually appropriate for counts, the improved linear model performed better for prediction in this case.

## Critical Reflection

I think the final model is good enough to be useful for Indego, but not good enough to fully replace human decision-making. A final MAE of about 0.675 trips per station-hour is a clear improvement over the earlier models, which means the model is learning important demand patterns. Still, the model performs worse during PM rush and in the busiest stations, which is when prediction errors are most costly for rebalancing. If the model underpredicts demand at those times, stations may run out of bikes or docks when they are needed most. For that reason, I would recommend using this system as a support tool for operations staff, especially for identifying likely demand patterns, but not as a fully automatic deployment system.

I also think there are real equity concerns. The errors are not random: they are clustered in some parts of the city and appear to vary with demographic characteristics. That means some neighborhoods may receive less accurate demand forecasts than others, which could translate into less reliable bike access. If that happens, the system could reinforce existing inequalities instead of improving service fairly across the city. To prevent this, Indego should evaluate model performance by neighborhood, not only citywide, and should set safeguards such as regular fairness audits and manual review for high-error areas. The model is also limited because it misses factors like holidays, school schedules, special events, and real-time disruptions. With more time and better data, I would add those features and also incorporate more detailed information about station capacity and rebalancing operations.
